Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-07-07 fstats_pca_lewontin.tex
submission

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\normsq}[1]{\left\lVert#1\right\rVert^2}
\newcommand{\MX}{\mathbf{X}} %uncentered data
\newcommand{\MC}{\mathbf{C}} %centering
\newcommand{\MY}{\mathbf{Y}} %centered data
\newcommand{\MF}{\mathbf{F}_2} %F2-distance matrix
\newcommand{\MFT}{\mathbf{F}_3} %F3-distance matrix
\newcommand{\MP}{\mathbf{P}} % PCs
\newcommand{\ML}{\mathbf{L}} % loadings
\newcommand{\MK}{\mathbf{K}} % Kernel
\newcommand{\MSINGULAR}{\mathbf{\Sigma}} % Singular values matrix
\newcommand{\MEIGEN}{\mathbf{\Lambda}} % Eigenvalue matrix
\newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel
\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
The genetic diversity in human population is shaped by our complex history. $F$-statistics are a widely used tool to model genetic diversity when populations are related acording to a graph and gene flow is rare. Here, I use a geometric interpretation of  to interpret $F$-statistic in the context of Principal Component Analysis (PCA), showing that the two approaches are closely related. 
I show that orthogonal projections are a key concept to establish this link, and that it leads to simple geometric interpretations of $F$-statistics on a PCA-plot. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that population structure is sparse, and only a few components contribute to most statistics, and I  develop novel visualizations that allow for checking the assumptions of more sophisticated models. My results allows interpreting  $F$-statistics in arbitrarily structured populations, to obtain  more complete and less biased models of human genetic variation.
\end{abstract}
\section{Introduction}
About 15\% of genetic variation in humans can be explained by population structure \citep{dobzhansky1972, barbujani1997, rosenberg2002a}, but the information contained in these 15\% is sufficient to study our genetic diversity and history in great detail \citep{cavalli-sforza1994,reich2018a}. For some data sets it is possible to predict an individuals origin at a resolution of a few hundred kilometers \citep{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.

Human genetic diversity is caused by a number of forces, and has both discrete and continuous components \citep{rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a}. Spatially isolated populations are expected to slowly diverge, particularly if they are separated by barriers to migration such as mountain ranges, oceans or deserts \citep{bradburd2013, peter2020a, rosenberg2005}. On the other hand, major population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity \citep{cavalli-sforza1994, ramachandran2005, novembre2008, peter2020a, stoneking2016, racimo2020}. Local migration between neighboring populations will flatten differentiation, and long-distance migrations \citep{alves2016} or secondary contact and admixture between diverged populations, such as Neandertals and early modern humans \citep{green2010} adds further complexity.

This complex population structure is frequently handled by using multiple models with different assumptions; each emphasizing a particular aspects of the data. Data-driven methods such as Principal Component Analysis (PCA) \citep{cavalli-sforza1994} structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are often used to display the full complexity of the data, but they have the disadvantage that they are not easily interpretable. For this purpose, more explicit demographic models \citep{gutenkunst2009, kamm2015, excoffier2013} are applied, which allow for parameter estimation or hypothesis tests.

Particularly in the analysis of human ancient DNA, a set of techniques based on $F$-statistics \textit{sensu} Patterson have risen in popularity \citep{patterson2012, peter2016}. This framework is based on the assumption that the relationship between three or four populations is often tree-like, and allows for a variety of tests of treeness and more complex demographic models \citep{patterson2012, harney2021}. However, the connections between PCA, $F$-statistics and demographic models are currently unclear, which makes quantitative comparisons, detecting model violations and joint interpretation of the results of these approaches difficult. Since both $F$-statistics and PCA are functions of expected pairwise coalescent times \citep{mcvean2009, peter2016}, this is one avenue to link these approaches. Here, I instead use the geometric interpretation of $F$-statistics introduced by  \cite{oteo-garcia2021} to directly visualize $F$-statsitics on a PCA plot.

\section{Theory}
In this section, I will give a very brief formal introduction to $F$-statistics and PCA. A more detailed technical treatise of PCA is given in e.g. \cite{jolliffe2013}, and a useful guide to interpretation is \cite{cavalli-sforza1994}. Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2020a} or \cite{oteo-garcia2021} helpful.

\subsection{Introduction to PCA}
Let us assume we have some genotype data summarized in a matrix $\MX$ whose entry $x_{ij}$ reflects the allele frequency of the $i$-th population at the $j$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. As a population may be represented by just one individual, there is no conceptual difference between these cases and I will
refer to populations as unit for analysis. Since the allele
frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}.

@@

\begin{comment}
	\begin{figure}
		\includegraphics[width=\textwidth]{pca_explanation.png}
		\caption{Basic Idea of PCA from 2D to 1D representation. A: Allele frequencies from different populations (blue dots) at two SNPs. A PCA is performed by centering the data (B), and rotating it (B) such that the first PC explains the majority of variation in the data, and the second PC is orthognal to the first, and explains the residual. A lower-dimensional approximation (in this case 1D) can be achieved by just keeping the first PC (E); which can be translated back to the original data space by inverting the rotation and centering (F).}
		\label{fig:pca_explanation}
	\end{figure}
\end{comment}

@@

	where $\MC = \mathbf{I} -\frac{1}{n}\mathbf{1}$ is a centering matrix that subtracts row means, with $\mathbf{I}, \mathbf{1}$  the identity matrix and a matrix of ones, respectively. The orthogonal matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and is used to reveal population structure. The SNP loadings $\ML=\mathbf{V}^T$ are an orthonormal matrix of size $n \times k$, its rows give the contribution of each SNP to each PC, it is often useful to look for outliers that might be indicative of selection \citep[e.g][]{francois2010}.

@@

\subsection{Introduction to $F$-statistics}
PCA is typically used to model population structure between many populations. $F$-statistics take the opposite approach, revealing the relationship between just two,  three or four populations at a time. The three $F$-statistics can be defined as 
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) \times S&=& \sum_{l=1}^S(x_{1l} - x_{2l})^2 &=& \normsq{X_1, X_2}\\
	&F_3(X_1; X_2, X_3) \times S&=& \sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) \times S&=& \sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. The number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently. Furthermore, both $F_3$ and $F_4$ can be written as sums of $F_2$-statistics:
\begin{subequations}
	\begin{align}
2F_3(X_1; X_2, X_3) &= 2F_2(X_2, X_3) - F_2(X_1, X_2) + F_2(X_1, X_3)\\
2F_4(X_1, X_2; X_3, X_4) &= F_2(X_1, X_3) + F_2(X_2, X_4) - F_2(X_1,X_4) - F_2(X_2, X_3)
	\end{align}
\end{subequations}

$F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}. In a tree, the squared Euclidean distance $F_2(X, Y)$ measures the length of all branches between populations $X$ and $Y$; and $F_3$ and $F_4$ represent external and internal branches in a tree, respectively \citep{peter2016}. The length is a measure of genetic drift, and is non-negative if data is generated under a tree \citep{patterson2012}. This interpretation is useful to understand a number of applications. The outgroup-$F_3$-statistic $F_3(O; U, X_i)$, for example, is useful if we have an unknown population $U$, and want to find its closest relatives from a panel of populations $X_i$. The highest values of $F_3$ indicate the population $X_i$ most closely related to $U$, using the outgroup $O$ to correct for differences in sample times. The population $X_i$ with the largest value is the most closely related population out of the reference sample. The internal branches described by  $F_4$-statistics are frequently used for complex models, such as  reconstructing admixture graphs \citep{patterson2012, lipson2013} and estimating admixture proportions \citep{petr2019, harney2021}.

@@

\subsection{Connection between PCA and $F$-statsitics}	
\subsubsection{Principal components from $F$-statistics}
PCA and $F$-statistics are closely related. In fact, the principal components can be directly calculated from $F$-statistics using multidimensional scaling \citep{gower1966}. Suppose we calculate the pairwise $F_2(X_i, X_j)$ between all $n$ populations, and collect them in a matrix $\MF$. We can obtain the principal components from this matrix by double-centering it, so that its row and column means are zero, and perform an eigendecomposition of the resulting matrix:
\begin{equation}
\MP\MP^T = - \frac{1}{2}\MC\MF\MC \text{.} \label{eq:mds}
\end{equation}

@@

\begin{align}
F_2(X_i, X_j) &=&  \sum_{l=1}^L \big( x_{il} -x_{jl}\big)^2  &&\nonumber\\ 
&=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2   &=& F_2(Y_i, Y_j) \nonumber\\
&=& \sum_{k=1}^n (p_{ik} - p_{jk})^2  &=& F_2(P_i, P_j) \text{,}\label{eq:fpc}
\end{align}
A detailed derivation of this is given in Appendix \ref{appendix:fonpc}.
As $F_3$ and $F_4$ can be written as sums of $F_2$-terms, analogous relations apply.

In most applications, we do not use all PCs, but instead use only the first $K$ PCs.
Thus, 
\begin{equation}
F_2(P_i, P_j) = \underbrace{\sum_{k=1}^K(p_{ik} - p_{jk})^2}_{\hat{F_2}^{(K)}(P_i, P_j)} + \sum_{k=K+1}^n(p_{ik} - p_{jk})^2 \text{.}
\end{equation}
If we sum up the approximation errors $F_2 - \hat F_2$ over all pairs of populations, we obtain the Frobenius-norm of the error $\normsq{\MF - \hat{\MF}}_F$; which is precisely the function that is minimized in MDS \citep{jolliffe2013}. Thus, $\hat\MF$ is the optimal approximation of $\MF$ for any $K$.

\section{Material \& Methods}
The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation on PCA plots, and can be approximated using PCA. I will explore this in detail in the next section, and illustrate it on two sample data sets that I briefly introduce here. Both are based on the analyses by \cite{lazaridis2014} and use the ``Human Origins''-SNP set (597,573 SNPs), based on the subset of the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}. The code used to create all figures and analyses will be available on [Archive to be created before publishing].

\paragraph{World Overview data set}
This data set represents global human genetic variation (638 individuals from 33 population), as used by \citep{lazaridis2014}. As this data set is very sparse, it may be well-approximated by an admixture graph.

\paragraph{West-Eurasian data set}
This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe. It is frequently used as a basis of comparison for ancient genetic analyses of Western Eurasian individuals \citep{patterson2012, lazaridis2014}. Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}, and thus not particularly graph-like.

\paragraph{Computation $F$-statistics and PCA}
I perform analyses at the level of populations to ease presentation.  It is an assumption of $F$-statistics that the genetic variation within sampled population is independent of the variation between samples. All computations are performed in \texttt{R}. I use \texttt{admixtools 2.0.0} \url{https://github.com/uqrmaie1/admixtools} to compute $F$-statistics. To obtain a PC-decomposition, I first calculate all pairwise $F_2$-statistics, obtain a nearby negative semidefinite matrix using the \texttt{nearPD} function, and then use equation \ref{eq:mds} and the \texttt{eigen} function to obtain the PCs.

@@

 Thus if $\hat{F}_3 \ll F_3$, the true circle will be bigger than would be predicted from a 2D-plot. In this case, substantial relevant genetic differentiation is ``hidden'' in the higher PCs. If this is the case, populations that appear inside the circle on a PCA-plot may, in fact, have positive $F_3$-statistics, because they are outside the $n$-ball in higher dimensions. However, if a population lies outside the circle on any 2D-projection, $F_3$ is guaranteed to be bigger than 0.

\paragraph{Example}
I visualize this on the admixture statistic $F_3(X; Basque, Turkish)$, on the first two PCs of the Westeurasian data set (Figure \ref{fig:f3}A). In this case, the projected $n$-ball (light gray) and circle based on 2D (dark gray) align relatively closely, but several populations inside the ball (e.g. Sardinian, Finnish) have, in fact, positive $F_3$-values, which reveals that the first two PCs do not capture all the genetic variation relevant for Southern European population structure. This is expected because for spatially continuous populations, PCA will not be sparse \citep{novembre2008a}.  Consequently, approximating $F_3$ by the first two or ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.

@@

\subsection{$F_3$ as a projection}
The inner product $\langle X_x - X_1, X_x - X_2 \rangle$ is closely related to the projection of one vector onto the other. This interpretation is useful either when calculating an outgroup-$F_3$-statistic, or when we want to find the ``best'' admixture source $X_2$ if we assume the admixed population $X_x$ and one source $X_1$ are known.  The angle between $X_x - X_1$ and $X_x - X_2$ is obtuse if $X_2$ is in the half-plane whose boundary goes through $X_x$ and is orthogonal to $\overline{X_XX_1}$ (Figure \ref{fig:geom}C), and its value is proportional to the projection onto $X_X - X_1$. A common interpretation guide is to use the population which yields the most negative $F_3$-statistic as the population which is most closely related to the ``true'' source of admixture \citep{patterson2012}; on a PCA-plot this corresponds to using the population that is furthest away from $X_x$ on the direction $X_x - X_1$.

@@

\paragraph{Example}
Again, these projection will be orthogonal when using the full data, and may only be approximately orthogonal when just using the first two PCs. In Figure \ref{fig:f3}D, I visualize this on the outgroup-$F_3$-statistic $F_3(Mbuti; Mozabite, X_i)$, i.e. a statistic that aims to find the population most closely related to Mozabite (a Berber ethnic group from the northern Sahara), assuming the Mbuti are an outgroup. On a PCA, we can interpret this $F_3$ statistic as the projection of the line segment from $\text{Mbuti}$ to population $X_i$ onto the line through Mbuti and Mozabite (black line). For each population, the projection is indicated with a grey line. In the full data space, this line is always orthogonal to the segment Mbuti-Mozabite, but on the plot (i.e.) the subspace spanned by the first two PCs, this is not necessarily the case. We see that particularly the samples from East Asia, Siberia and the Americas project very close to orthogonally, suggesting that most of the variation is captured by these first two PCs, and the coloring based on the full $F_3$-statistics shows that in this case, the first two PCs approximate the $F_3$-statistic very well. We can quantify this and find that the first two PCs slightly underestimate the absolute value of $F_3$ (Figure \ref{fig:f3}E), but keep the relative ordering. This $F_3$-statistic is also very sparse, with e.g. PCs 3-5, 7 and 10 having almost zero contribution to all statistics (Figure \ref{fig:f3}F), and PCs 6, 8 and 9 having a similar non-zero contribution for almost all statistics, likely because these PCs explain within-African variation.

\subsection{$F_4$-statisticss as angles}
The interpretation of $F_4$ in PCA is similar to that of $F_3$ as a projection of one vector onto another, with the difference that now all four points may be distinct. As for $F_3$, a finding of $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3$ and $X_4$ are orthogonal, or that the two populations map to the same point, and otherwise it will correspond to the length of the projection (Figure \ref{fig:geom}E).

>>>>

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\normsq}[1]{\left\lVert#1\right\rVert^2}
\newcommand{\MX}{\mathbf{X}} %uncentered data
\newcommand{\MC}{\mathbf{C}} %centering
\newcommand{\MY}{\mathbf{Y}} %centered data
\newcommand{\MF}{\mathbf{F}_2} %F2-distance matrix
\newcommand{\MFT}{\mathbf{F}_3} %F3-distance matrix
\newcommand{\MP}{\mathbf{P}} % PCs
\newcommand{\ML}{\mathbf{L}} % loadings
\newcommand{\MK}{\mathbf{K}} % Kernel
\newcommand{\MSINGULAR}{\mathbf{\Sigma}} % Singular values matrix
\newcommand{\MEIGEN}{\mathbf{\Lambda}} % Eigenvalue matrix
\newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel
\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.
\end{abstract}
\section{Introduction}
Most genetic variation in humans is shared between all of us, but around 15\% of genetic variation in humans can be explained by population structure \citep{dobzhansky1972, barbujani1997, rosenberg2002a}. Genome-scale data has allowed us to leverage the information contained in these 15\% to study our genetic diversity and history in great detail \citep{cavalli-sforza1994,reich2018a}. For some data sets it is possible to predict an individuals origin at a resolution of a few hundred kilometers \citep{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.

However, understanding, conceptualizing and modeling this variation is far from trivial, particularly since misconstrued models of human genetic differentiation have repeatedly been used to justify racist, eugenic and genocidal policies. Lewontin's landmark \citeyear{dobzhansky1972} paper on the apportionment of human genetic diversity was one of the first to question ``races'' not only on ethical, but also on scientific grounds. In Lewontin's data, less than half of the already small proportion of between-population genetic variation can be attributed to ``racial'' continental-scale groupings. Over the last five decades, this view has been corroborated and extended many times \citep{cann1987, cavalli-sforza1994, cann2002, rosenberg2002, reich2018a}, and we now know that human races are an arbitrary, biologically useless polyphyletic grouping that is maintained by historical and social conventions.

Modern descriptions of human genetic diversity focus on the evolutionary processes that caused it, and which gives rise to  both discrete and continuous components \citep{rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a}. In isolation, populations are expected to slowly diverge, particularly if they are separated by barriers to migration such as mountain ranges, oceans or deserts \citep{bradburd2013, peter2020a, rosenberg2005}. On the other hand, major population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity \citep{cavalli-sforza1994, ramachandran2005, novembre2008, peter2020a, stoneking2016, racimo2020}. Local migration between neighboring populations will flatten differentiation, and long-distance migrations \citep{alves2016} or secondary contact and admixture between diverged populations, such as Neandertals and modern humans \citep{green2010},will lead to locally increased diversity.

This complex population structure is frequently handled by using multiple models with different assumptions; each emphasizing a particular aspects of the data. Data-driven methods such as Principal Component Analysis (PCA, \cite{cavalli-sforza1994}) structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are often used to display the full complexity of the data, but they have the disadvantage that they are not easily interpretable. For this purpose, more explicit demographic models \citep{gutenkunst2009, kamm2015, excoffier2013} are applied, which allow for parameter estimation or hypothesis tests.

Particularly in the analysis of human ancient DNA, a set of techniques based on $F$-statistics \textit{sensu} Patterson have risen in popularity \citep{patterson2012, peter2016}. This framework is based on the null-assumption that the relationship between  populations is tree-like, and then uses tests involving three or four populations to probe for violations of treeness \citep{patterson2012}. In a further step, $F$-statistics of multiple populations can be combined to estimate parameters in more complex models \citep{patterson2012, harney2021}. However, the connections between PCA, $F$-statistics and explicit demographic models are often unclear, which makes quantitative comparisons, detecting model violations and joint interpretation of these approaches difficult. Since both $F$-statistics and PCA are functions of expected pairwise coalescent times \citep{mcvean2009, peter2016}, this is one avenue to link these approaches. Here, I instead use the geometric interpretation of $F$-statistics developed by  \cite{oteo-garcia2021} to directly visualize $F$-statistics on a PCA plot.

\section{Theory}
In this section, I will give a very brief formal introduction to $F$-statistics and PCA. A more detailed technical treatise of PCA is given in e.g. \cite{jolliffe2013}, and a useful guide to interpretation is \cite{cavalli-sforza1994}. Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2016} or \cite{oteo-garcia2021} helpful.

\subsection{Introduction to PCA}
Let us assume we have some genotype data summarized in a matrix $\MX$ whose entry $x_{ij}$ reflects the allele frequency of the $i$-th population at the $j$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. As a population may be represented by just one individual, there is no conceptual difference between an individual-based and population-based analysis. Since the allele
frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}.

@@

@@

	where $\MC = \mathbf{I} -\frac{1}{n}\mathbf{1}$ is a centering matrix that subtracts row means, with $\mathbf{I}, \mathbf{1}$  the identity matrix and a matrix of ones, respectively. The orthogonal matrix of principal components $\MP=\mathbf{U}\MSINGULAR$ has size $n \times n$ and is used to reveal population structure. The SNP loadings $\ML=\mathbf{V}^T$ are an orthonormal matrix of size $n \times S$, its rows give the contribution of each SNP to each PC, it is often useful to look for outliers that might be indicative of selection \citep[e.g][]{francois2010}.

@@

\subsection{Introduction to $F$-statistics}
PCA is typically used to model population structure between many populations. $F$-statistics take the opposite approach, revealing the relationship between just two,  three or four populations at a time. The three $F$-statistics can be defined as 
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2 &=& \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. The normalization by the number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
\begin{align}
\langle a, b \rangle &= \sum_i a_ib_i\\
\langle a, b \rangle &= \norm{a}\norm{b}\cos{\phi}\\
\langle a, a \rangle &= \normsq{a}\\
\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$.

 Furthermore, both $F_3$ and $F_4$ can be written as sums of $F_2$-statistics:
\begin{subequations}
	\begin{align}
2F_3(X_1; X_2, X_3) &= F_2(X_2, X_3) - F_2(X_1, X_2) - F_2(X_1, X_3)\label{eq:f3fromf2}\\
2F_4(X_1, X_2; X_3, X_4) &= F_2(X_1, X_3) + F_2(X_2, X_4) - F_2(X_1,X_4) - F_2(X_2, X_3)\label{eq:f4fromf2}
	\end{align}
\end{subequations}

$F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}. In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of all branches between populations $X_1$ and $X_2$; and $F_3$ and $F_4$ represent external and internal branches in a tree, respectively \citep{peter2016}. The length is a measure of genetic drift, and is non-negative if data is generated under a tree \citep{patterson2012}. This interpretation is useful to understand a number of applications. The outgroup-$F_3$-statistic $F_3(O; U, X_i)$, for example, is useful if we have an unknown population $U$, and want to find its closest relatives from a panel of populations $X_i$. The highest values of $F_3$ indicate the population $X_i$ most closely related to $U$, using the outgroup $O$ to correct for differences in sample times. The population $X_i$ with the largest value is the most closely related population out of the reference sample. The internal branches described by  $F_4$-statistics are frequently used for complex models, such as  reconstructing admixture graphs \citep{patterson2012, lipson2013} and estimating admixture proportions \citep{petr2019, harney2021}.

@@

\subsection{Connection between PCA and $F$-statsitics}	
\subsubsection{Principal components from $F$-statistics}
PCA and $F$-statistics are closely related. In fact, the principal components can be directly calculated from $F$-statistics using multidimensional scaling, which, for squared Euclidean ($F_2$)-distances, leads to an identical decomposition to PCA \citep{gower1966}. Suppose we calculate the pairwise $F_2(X_i, X_j)$ between all $n$ populations, and collect them in a matrix $\MF$. We can obtain the principal components from this matrix by double-centering it, so that its row and column means are zero, and perform an eigendecomposition of the resulting matrix:
\begin{equation}
\MP\MP^T = - \frac{1}{2}\MC\MF\MC \text{.} \label{eq:mds}
\end{equation}

@@

\begin{align}
F_2(X_i, X_j) &=&  \sum_{l=1}^L \big( x_{il} -x_{jl}\big)^2  &&\nonumber\\ 
&=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2   &=& F_2(Y_i, Y_j) \nonumber\\
&=& \sum_{k=1}^n (p_{ik} - p_{jk})^2  &=& F_2(P_i, P_j) \text{,}\label{eq:fpc}
\end{align}
A derivation of this change-of-basis is given in Appendix \ref{appendix:fonpc}.
As $F_3$ and $F_4$ can be written as sums of $F_2$-terms (Eqs. \ref{eq:f3fromf2}, \ref{eq:f4fromf2}), analogous relations apply.

In most applications, we do not use all PCs, but instead truncate to the first $K$ PCs, which explain most of the between-population genetic variation.
Thus, 
\begin{equation}
F_2(P_i, P_j) = \underbrace{\sum_{k=1}^K(p_{ik} - p_{jk})^2}_{\hat{F_2}^{(K)}(P_i, P_j)} + \sum_{k=K+1}^n(p_{ik} - p_{jk})^2 \text{.}
\end{equation}
If we sum up the approximation errors $F_2 - \hat F_2$ over all pairs of populations, we obtain the Frobenius-norm of the error $\normsq{\MF - \hat{\MF}}_F$; which is precisely the function that is minimized in MDS \citep{jolliffe2013}. Thus, $\hat\MF^{(K)}$ is an optimal sparse approximation of $\MF$ for any $K$.

\section{Material \& Methods}
The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation in PCA-space, which can be approximated on PCA plots. In the next section I explore this connection in detail, and illustrate it on two sample data sets that I briefly introduce here. Both are based on the analyses by \cite{lazaridis2014}. The data is from the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}, using data on the ``Human Origins''-SNP set (597,573 SNPs). SNPs with missing data in any population are excluded. The code used to create all figures and analyses will be available on [Archive to be created before publishing].

\paragraph{``World'' data set}
This data set represents global human genetic variation (638 individuals from 33 population), as used by \citep{lazaridis2014}. As this data set is very sparse, it may be well-approximated by an admixture graph.

\paragraph{West-Eurasian data set}
This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe. It is frequently used as a basis of comparison for ancient genetic analyses of Western Eurasian individuals \citep{patterson2012, lazaridis2014}. Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}, and thus may not be particularly graph-like.

\paragraph{Computation $F$-statistics and PCA}
I perform analyses at the level of populations to ease presentation.  It is an assumption of $F$-statistics that the genetic variation within sampled population is independent of the variation between samples \citep{patterson2012}. All computations are performed in \texttt{R}. I use \texttt{admixtools 2.0.0} \url{https://github.com/uqrmaie1/admixtools} to compute $F$-statistics. To obtain a PC-decomposition, I first calculate all pairwise $F_2$-statistics, obtain a nearby negative semidefinite matrix using the \texttt{nearPD} function, and then use equation \ref{eq:mds} and the \texttt{eigen} function to obtain the PCs.

@@

 Thus if $\hat{F}_3 \ll F_3$, the true circle will be bigger than would be predicted from a 2D-plot. In this case, substantial relevant genetic differentiation is ``hidden'' in the higher PCs, and populations that appear inside the circle on a PCA-plot may, in fact, have positive $F_3$-statistics. This is because they are outside the $n$-ball in higher dimensions. The converse interpretation is more strict: if a population lies outside the circle on \emph{any} 2D-projection, $F_3$ is guaranteed to be bigger than 0.

\paragraph{Example}
As an example, I visualize the admixture statistic $F_3(X; \text{Basque}, \text{Turkish})$, on the first two PCs of the Westeurasian data set (Figure \ref{fig:f3}A). In this case, the projected $n$-ball (light gray) and circle based on 2D (dark gray) align relatively closely, but several populations inside the ball (e.g. Sardinian, Finnish) have, in fact, positive $F_3$-values. This reveals that the first two PCs do not capture all the genetic variation relevant for Southern European population structure. This is expected because for spatially continuous populations, PCA will not be sparse \citep{novembre2008a}.  Consequently, approximating $F_3$ by the first two or ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.

@@

\subsection{$F$-statistics as projections}
The inner product $\langle X_x - X_1, X_x - X_2 \rangle$ is closely related to the projection of a vector onto another one. This interpretation is useful either when calculating an outgroup-$F_3$-statistic, or when we want to find the ``best'' admixture source $X_2$ if we assume the admixed population $X_x$ and one source $X_1$ are known.  The angle between $X_x - X_1$ and $X_x - X_2$ is obtuse if $X_2$ is in the half-plane whose boundary goes through $X_x$ and is orthogonal to $\overline{X_XX_1}$ (Figure \ref{fig:geom}C), and its value is proportional to the projection onto $X_X - X_1$. A common interpretation guide is to use the population which yields the most negative $F_3$-statistic as the population which is most closely related to the ``true'' source of admixture \citep{patterson2012}; on a PCA-plot this corresponds to using the population that is furthest away from $X_x$ on the direction $X_x - X_1$.

@@

\paragraph{Example}
Again, these projection will be orthogonal when using the full data, and may only be approximately orthogonal when approximated using the first two PCs. In Figure \ref{fig:f3}D, I visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Mozabite}, X_i)$, i.e. a statistic that aims to find the population most closely related to Mozabite (a Berber ethnic group from the northern Sahara), assuming the Mbuti are an outgroup. On a PCA, we can interpret this $F_3$ statistic as the projection of the line segment from $\text{Mbuti}$ to population $X_i$ onto the line through Mbuti and Mozabite (black line). For each population, the projection is indicated with a grey line. In the full data space, this line is always orthogonal to the segment Mbuti-Mozabite, but on the plot (i.e. the subspace spanned by the first two PCs), this is not necessarily the case. We see that particularly the samples from East Asia, Siberia and the Americas project very close to orthogonally, suggesting that most of the variation is captured by these first two PCs, and the coloring based on the full $F_3$-statistics shows that in this case, the first two PCs approximate the $F_3$-statistic very well. We can quantify this and find that the first two PCs slightly underestimate the absolute value of $F_3$ (Figure \ref{fig:f3}E), but keep the relative ordering. This $F_3$-statistic is also very sparse, with e.g. PCs 3-5, 7 and 10 having almost zero contribution to all statistics (Figure \ref{fig:f3}F), and PCs 6, 8 and 9 having a similar non-zero contribution for almost all statistics, likely because these PCs explain within-African variation.

\subsection{$F_4$-statistics as angles}
The interpretation of $F_4$ in PCA is similar to that of $F_3$ as a projection of one vector onto another, with the difference that now all four points may be distinct. As for $F_3$, a finding of $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal, or that the two populations map to the same point, and otherwise it will correspond to the length of the projection (Figure \ref{fig:geom}E).
